---
title: "HW1"
author: "Namit Shrivastava"
format: pdf
---

## Study Objectives

1. On average, how does systolic blood pressure change over time as people get older?
2. How much variance among participants is there in the trajectories of systolic blood pressure, and can this variance be explained by sex and age?

# Data Preparation and Loading

```{r}
library(tidyverse)
library(lme4)
library(broom.mixed)
library(knitr)
library(kableExtra)
library(ggplot2)
library(performance)
```

```{r}
# Loading the data
load("/Users/namomac/Desktop/SURV-687/hrs_wbbp_long.rda")

# Examining the structure of the data
str(hrs_wbbp_long)
head(hrs_wbbp_long)

# Checking for missing values
colSums(is.na(hrs_wbbp_long))
```

```{r}
# Creating factor variables
hrs_wbbp_long <- hrs_wbbp_long %>%
  mutate(
    RAGENDER_f = factor(RAGENDER, levels = c(1, 2), labels = c("Male", "Female")),
    HHIDPN_f = factor(HHIDPN),
    wave_f = factor(wave)
  )

# Checking the range of key variables
summary(hrs_wbbp_long)
```

Ok so based on this dataset of 1,555 blood pressure observations from 311 war babies cohort participants, I see there are no missing values and I can see that each person was measured up to every four years from year 0 to year 16 with systolic blood pressure ranging from about 79 to 216 mmHg with a typical value around 126 mmHg and the participants’ ages span roughly 30 to 92 years with a median near 67, and about 44% are male and 56% are female.

Now I can also see that the “year” is evenly spaced with quartiles at 0, 4, 8, 12, and a maximum of 16, reflecting the 4-year follow-up design i.e. waves are coded as 8, 10, 12, 14, and 16, each appearing 311 times, which is consistent with five repeated measurement occasions for many participants.

The blood pressure distribution centers near 125–127 mmHg with interquartile range roughly 115 to 138 mmHg, which suggests moderate variability across people and occasions and hence, ages are concentrated in the 60s to early 70s (median 67, IQR 62–72), which fits the cohort profile.

Now, what I also observed was that the gender coding shows 685 male and 870 female observations (i.e. in baseline counts these correspond to participant-level sex), and the ID field HHIDPN has many repeated entries (commonly 5 per person), and hence confirming the longitudinal structure required for multilevel modeling in later steps.

# Question 1: Describe the sample in a reasonable way, i.e., construct a table or a graph of descriptive statistics as you may do for a publication. Be sure to remember that there is more than one observation per subject. Comment briefly on the general characteristics of the observations. Keep the study objectives in mind with your descriptive presentation.

```{r}
# Showcasing sample size information
n_participants <- length(unique(hrs_wbbp_long$HHIDPN))
n_observations <- nrow(hrs_wbbp_long)

cat("Total number of participants:", n_participants, "\n")
cat("Total number of observations:", n_observations, "\n")
cat("Average observations per participant:", round(n_observations/n_participants, 2), "\n")

# Now to check observations per participant
obs_per_person <- hrs_wbbp_long %>%
  group_by(HHIDPN) %>%
  summarise(n_obs = n())

table(obs_per_person$n_obs)
```

So, based on these results, I can see that the dataset has a perfectly balanced longitudinal structure. Here, I have 311 participants from the War Babies cohort, and each person has exactly 5 blood pressure measurements over time, and hence, giving me a total of 1,555 observations. This means there's no missing data whatsoever and that every single participant was measured at all five time points (waves 8, 10, 12, 14, and 16), which is quite remarkable for a longitudinal study spanning 16 years. 

Now this complete balanced design is actually ideal for multilevel modeling because one doe'st have to worry about missing data patterns or unequal numbers of observations per person affecting this analysis. The fact that everyone has exactly 5 observations means this statistical models will have equal precision for estimating individual trajectories, and I can confidently examine how systolic blood pressure changes over time both at the population level and for individual participants.


```{r}
# Now to explore the baseline characteristics i.e. first 
# observation per person in wave 8
baseline_chars <- hrs_wbbp_long %>%
  group_by(HHIDPN) %>%
  slice_min(as.numeric(wave)) %>%
  ungroup()

# Creating descriptive statistics table
desc_stats <- baseline_chars %>%
  summarise(
    n = n(),
    age_mean = round(mean(age_y, na.rm = TRUE), 1),
    age_sd = round(sd(age_y, na.rm = TRUE), 1),
    age_range = paste(round(min(age_y), 1), "-", round(max(age_y), 1)),
    bp_sys_mean = round(mean(bp_sys, na.rm = TRUE), 1),
    bp_sys_sd = round(sd(bp_sys, na.rm = TRUE), 1),
    bp_sys_range = paste(round(min(bp_sys), 1), "-", round(max(bp_sys), 1))
  )
```
```{r}
# Gender distribution at participant level as mentioned in the question
gender_dist <- baseline_chars %>%
  group_by(RAGENDER_f) %>%
  summarise(n = n(), percentage = round(n()/nrow(baseline_chars)*100, 1))

kable(gender_dist, caption = "Gender Distribution at Baseline (N = 311 participants)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

kable(desc_stats, 
      col.names = c("N", "Age Mean", "Age SD", "Age Range", "BP Sys Mean", "BP Sys SD", "BP Sys Range"),
      caption = "Baseline Characteristics (Wave 8)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

So based on my baseline characteristics analysis above, I can see that the sample is well-balanced and representative of the War Babies cohort. 

At the first measurement in wave 8, I have 311 participants with a slight female majority i.e., 174 women (55.9%) and 137 men (44.1%). I can assume that the participants were relatively young when the study began, with an average age of 58.8 years and ages ranging from 30 to 76 years, it shows that while most were in their late 50s and early 60s, there's good representation across different ages within this birth cohort. 

Looking at their blood pressure at baseline, I find that the average systolic blood pressure was 124.5 mmHg, which is right at the upper end of the normal range based on clinical guidelines. So, there's considerable variation in blood pressure levels among participants, with a standard deviation of 19 mmHg and values ranging from a low of 83 mmHg to a high of 192 mmHg. 

I feel that this wide range suggests, I have participants spanning from those with very healthy blood pressure to those with moderate to severe hypertension at study entry. This baseline variability is actually quite valuable for such analysis because it means I can examine how blood pressure changes over time across people starting from very different health statuses.

Hence, this will help me better understand individual differences in trajectories and the factors that might influence them.

```{r}
# Longitudinal descriptive statistics by wave
wave_stats <- hrs_wbbp_long %>%
  group_by(wave) %>%
  summarise(
    n = n(),
    age_mean = round(mean(age_y, na.rm = TRUE), 1),
    age_sd = round(sd(age_y, na.rm = TRUE), 1),
    bp_sys_mean = round(mean(bp_sys, na.rm = TRUE), 1),
    bp_sys_sd = round(sd(bp_sys, na.rm = TRUE), 1),
    year_mean = round(mean(year, na.rm = TRUE), 1)
  ) %>%
  arrange(as.numeric(wave))

kable(wave_stats, 
      col.names = c("Wave", "N", "Age Mean", "Age SD", "BP Sys Mean", "BP Sys SD", "Year"),
      caption = "Characteristics by HRS Wave") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

Interestingly, when I look at how characteristics change across the five waves of data collection, I can see some interesting longitudinal patterns. As expected, the participants age exactly 4 years between each wave, starting at an average of 58.8 years in wave 8 and reaching 74.8 years by wave 16, with remarkably consistent age standard deviations of 4.8 years across all waves. Hence, this consistency confirms the balanced nature of the data i.e. no one dropped out over the 16-year study period. 

And then, upon analysing the blood pressure patterns, I can see that they don't show a simple linear increase over time like I might have initially expected. The average systolic blood pressure starts at 124.5 mmHg in wave 8, rises to 127.6 mmHg by wave 10, but then appears to stabilize around 126-127 mmHg for the remaining waves. This suggests that while there's some increase in blood pressure as people move from their late 50s to their mid-70s, it's not a dramatic upward trend but rather, it seems to level off after the initial years. 

The standard deviations also decrease slightly over time (from 19.0 to around 17.0-17.5 mmHg), which might indicate that blood pressure becomes somewhat more homogeneous within the group as they age, possibly due to medical interventions or I guess natural biological processes. So, I believe this longitudinal pattern will be crucial for my multilevel modeling, as it suggests both individual variation in trajectories and potentially non-linear population-level changes that my models will need to capture.


```{r}
# Additional summary showing the longitudinal structure
cat("Data Structure Summary:\n")
cat("- Total participants:", length(unique(hrs_wbbp_long$HHIDPN)), "\n")
cat("- Total observations:", nrow(hrs_wbbp_long), "\n")
cat("- Observations per participant:", nrow(hrs_wbbp_long)/length(unique(hrs_wbbp_long$HHIDPN)), "\n")
cat("- Study duration:", max(hrs_wbbp_long$year), "years\n")
cat("- Measurement interval: Every 4 years\n")
cat("- HRS Waves included:", paste(sort(unique(hrs_wbbp_long$wave)), collapse = ", "), "\n")
```

So to summarize again, I see that I'm working with 311 participants who have been followed consistently over a 16-year period, with measurements taken every 4 years at HRS waves 8, 10, 12, 14, and 16. Here, I have exactly 5 observations per participant, totaling 1,555 observations, which means there's absolutely no missing data i.e. every single person was measured at all five time points. 

I think this kind of complete follow-up over such a long period is quite rare in longitudinal research and it actually speaks to the quality of the Health and Retirement Study. Now, the 4-year measurement intervals give me a good balance between capturing long-term changes while maintaining a manageable study design. 

Hence, this consistent structure means I can examine how blood pressure trajectories unfold over time from when participants were in their late 50s to their mid-70s, and I feel that I don't have to worry about missing data patterns complicating my multilevel models or creating bias in my results later on I guess.


```{r descriptive-plots}
# Plot 1: Individual trajectories
p1 <- ggplot(hrs_wbbp_long, aes(x = year, y = bp_sys, group = HHIDPN)) +
  geom_line(alpha = 0.3, color = "gray60") +
  geom_smooth(aes(group = 1), method = "lm", se = TRUE, color = "blue", size = 1.5) +
  labs(title = "Individual Systolic Blood Pressure Trajectories Over Time",
       x = "Years Since First Measurement",
       y = "Systolic Blood Pressure (mmHg)") +
  theme_minimal()

# Plot 2: By gender
p2 <- ggplot(hrs_wbbp_long, aes(x = year, y = bp_sys, color = RAGENDER_f)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "Systolic Blood Pressure Trajectories by Gender",
       x = "Years Since First Measurement",
       y = "Systolic Blood Pressure (mmHg)",
       color = "Gender") +
  theme_minimal()

# Plot 3: Distribution of blood pressure
p3 <- ggplot(hrs_wbbp_long, aes(x = bp_sys)) +
  geom_histogram(bins = 30, fill = "lightblue", alpha = 0.7) +
  labs(title = "Distribution of Systolic Blood Pressure",
       x = "Systolic Blood Pressure (mmHg)",
       y = "Frequency") +
  theme_minimal()

print(p1)
print(p2)
print(p3)
```

## Comment on Descriptive Statistics

The sample consists of 311 participants from the War Babies cohort with exactly 5 observations per person over the study period (1,555 total observations). The data spans waves 8, 10, 12, 14, and 16, measured every 4 years from baseline (year 0) to 16 years of follow-up. At baseline, participants ranged in age from 30 to 92 years (mean = 66.8 years) with systolic blood pressure ranging from 78.5 to 215.5 mmHg (mean = 126.5 mmHg). The sample contains 685 males (44.1%) and 870 females (55.9%). The data shows a complete balanced longitudinal design with no missing values, which is ideal for multilevel modeling. The wide age range and substantial variation in blood pressure (both within and between individuals) provide excellent opportunity to examine individual differences in trajectories.

# Question 2: Basic Multilevel Model

I'll fit a multilevel model with systolic blood pressure as the dependent variable and time (year) as the independent variable, accounting for within-subject correlation.

```{r basic-mlm}
# Fit the basic multilevel model
model1 <- lmer(bp_sys ~ year + (year | HHIDPN), data = hrs_wbbp_long)

# Display model summary
summary(model1)

# Get confidence intervals
confint(model1)

# Extract coefficients in a tidy format
model1_tidy <- tidy(model1, conf.int = TRUE)
kable(model1_tidy, digits = 3, caption = "Model 1: Basic Multilevel Model Results") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Reasoning for Model Choice

I chose a random intercept and random slope model because:

1. **Random Intercepts**: Individuals likely have different baseline systolic blood pressure levels due to genetic, lifestyle, and other unmeasured factors.

2. **Random Slopes**: The rate of change in blood pressure over time likely varies across individuals - some may experience steeper increases than others.

3. **Within-Subject Correlation**: The repeated measures on the same individuals are correlated, which this model structure explicitly accounts for.

This model allows both the intercept (baseline blood pressure) and the slope (rate of change over time) to vary randomly across participants.

# Question 3: Mathematical Notation

The multilevel model fitted in Question 2 can be written mathematically as:

**Level 1 (Within-Person) Model:**
$$BP_{ij} = \beta_{0i} + \beta_{1i} \cdot Year_{ij} + e_{ij}$$

**Level 2 (Between-Person) Model:**
$$\beta_{0i} = \gamma_{00} + u_{0i}$$
$$\beta_{1i} = \gamma_{10} + u_{1i}$$

**Combined Model:**
$$BP_{ij} = \gamma_{00} + \gamma_{10} \cdot Year_{ij} + u_{0i} + u_{1i} \cdot Year_{ij} + e_{ij}$$

Where:
- $BP_{ij}$ = Systolic blood pressure for person $i$ at time $j$
- $Year_{ij}$ = Years since first measurement for person $i$ at time $j$
- $\gamma_{00}$ = Fixed intercept (average baseline blood pressure)
- $\gamma_{10}$ = Fixed slope (average rate of change per year)
- $u_{0i}$ = Random intercept for person $i$ (deviation from average baseline)
- $u_{1i}$ = Random slope for person $i$ (deviation from average rate of change)
- $e_{ij}$ = Level-1 residual

**Assumptions:**
- $u_{0i}, u_{1i} \sim N(0, \Sigma_u)$ where $\Sigma_u = \begin{pmatrix} \sigma_{u0}^2 & \sigma_{u01} \\ \sigma_{u01} & \sigma_{u1}^2 \end{pmatrix}$
- $e_{ij} \sim N(0, \sigma_e^2)$

# Question 4: Interpretation of Model Parameters

```{r model1-interpretation}
# Extract variance components
vc1 <- as.data.frame(VarCorr(model1))
kable(vc1, digits = 3, caption = "Variance Components - Model 1") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Calculate ICC
icc_intercept <- vc1$vcov[1] / (vc1$vcov[1] + vc1$vcov[4])
cat("ICC for intercepts:", round(icc_intercept, 3), "\n")
```

## Interpretation with Respect to Research Objectives

**Fixed Effects:**

- **Intercept (γ₀₀)**: The average systolic blood pressure at the first measurement (year = 0, wave 8) is estimated from the model output. This represents the baseline blood pressure level.

- **Year Slope (γ₁₀)**: On average, systolic blood pressure changes by the estimated coefficient per year. This addresses research objective 1, showing how blood pressure changes with time/age.

**Random Effects (Variance Components):**

- **Random Intercept Variance (σ²ᵤ₀)**: The between-person variance in baseline blood pressure indicates individual differences in starting blood pressure levels.

- **Random Slope Variance (σ²ᵤ₁)**: The between-person variance in yearly blood pressure change addresses research objective 2, showing whether there is meaningful individual variation in how blood pressure changes over time.

- **Correlation between Random Effects**: The correlation between individual baseline levels and rates of change indicates whether individuals with higher/lower baseline blood pressure tend to have steeper/flatter trajectories.

- **Residual Variance (σ²ₑ)**: The within-person variance represents measurement error and short-term fluctuations.

**Intraclass Correlation (ICC)**: The proportion of total variance that is between individuals, indicating the degree of clustering within persons.

# Question 5: Testing Random Year Effects

## Steps to Test the Null Hypothesis

To test H₀: σ²ᵤ₁ = 0 (variance of random year effects equals zero), I would follow these steps:

```{r likelihood-ratio-test}
# Fit reduced model without random slopes
model1_reduced <- lmer(bp_sys ~ year + (1 | HHIDPN), data = hrs_wbbp_long)

# Perform likelihood ratio test
lr_test <- anova(model1_reduced, model1)
print(lr_test)

# Calculate the test statistic and p-value manually
chi_sq <- 2 * (logLik(model1) - logLik(model1_reduced))
p_value <- pchisq(chi_sq, df = 2, lower.tail = FALSE) / 2  # Divide by 2 for boundary test

cat("Chi-square test statistic:", round(as.numeric(chi_sq), 3), "\n")
cat("p-value (boundary corrected):", round(p_value, 6), "\n")
```

## Steps for Hypothesis Testing:

1. **Fit the full model** with random intercepts and slopes
2. **Fit the reduced model** with only random intercepts  
3. **Compute the likelihood ratio test statistic**: LR = 2(LogLik_full - LogLik_reduced)
4. **Apply boundary correction**: Since we're testing a variance component at the boundary (≥ 0), divide the p-value by 2
5. **Compare to critical value**: Use a mixture of χ² distributions

## Conclusion

The likelihood ratio test yields a highly significant result (p < 0.001), indicating strong evidence against the null hypothesis. We reject H₀ and conclude that there is significant between-person variance in the yearly rate of blood pressure change. This supports research objective 2, confirming that individuals indeed vary meaningfully in their blood pressure trajectories over time.

# Question 6: Model with Sex and Age Effects

```{r model-with-covariates}
# Center age for better interpretation
hrs_wbbp_long <- hrs_wbbp_long %>%
  mutate(age_centered = age_y - mean(age_y, na.rm = TRUE))

# Fit model with sex and age effects
model2 <- lmer(bp_sys ~ year + RAGENDER_f + age_centered + 
               (year | HHIDPN), data = hrs_wbbp_long)

summary(model2)

# Get tidy results
model2_tidy <- tidy(model2, conf.int = TRUE)
kable(model2_tidy, digits = 3, caption = "Model 2: Multilevel Model with Sex and Age") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Variance components
vc2 <- as.data.frame(VarCorr(model2))
kable(vc2, digits = 3, caption = "Variance Components - Model 2") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# Question 7: Interpretation of Model 2

## Practical Interpretation

**Fixed Effects:**

- **Intercept**: For a male participant at average age (~67 years) at the first measurement, the expected systolic blood pressure is given by the model intercept.

- **Year Effect**: Controlling for sex and age, systolic blood pressure changes by the estimated coefficient per year on average.

- **Gender Effect**: Females have, on average, a difference in systolic blood pressure compared to males (reference category), controlling for age and time.

- **Age Effect**: For each year older a participant is (at any given measurement), their systolic blood pressure is expected to differ by the estimated coefficient, controlling for sex and time trend.

**Random Effects:**
The inclusion of sex and age as covariates may reduce the between-person variance components, indicating whether these variables explain some of the individual differences in blood pressure trajectories.

# Question 8: Age Variable Recoding and Model Considerations

```{r age-recoding-exploration}
# Explore age recoding options
summary(hrs_wbbp_long$age_y)

# Option 1: Age at study entry (baseline age)
hrs_wbbp_long <- hrs_wbbp_long %>%
  group_by(HHIDPN) %>%
  mutate(baseline_age = first(age_y[order(wave)])) %>%
  ungroup() %>%
  mutate(baseline_age_centered = baseline_age - mean(baseline_age, na.rm = TRUE))

# Check correlation between year and age
cor(hrs_wbbp_long$year, hrs_wbbp_long$age_y)

# Option 2: Age at 60 (meaningful clinical threshold)
hrs_wbbp_long <- hrs_wbbp_long %>%
  mutate(age_at_60 = age_y - 60)

# Fit model with baseline age instead of time-varying age
model3 <- lmer(bp_sys ~ year + RAGENDER_f + baseline_age_centered + 
               (year | HHIDPN), data = hrs_wbbp_long)

summary(model3)
```

## Age Recoding Recommendations

**How to recode age for better interpretation:**

1. **Use baseline age instead of time-varying age**: Since age and year are perfectly correlated within individuals (r ≈ 1.0), including both creates multicollinearity issues.

2. **Center at a meaningful value**: I recommend centering baseline age at 60 years, which is:
   - Close to the sample mean
   - A meaningful clinical threshold for cardiovascular risk assessment
   - Easier to interpret than the sample mean

**Impact on interpretation:**
- The intercept would represent expected blood pressure for a 60-year-old male at baseline
- Age effects would represent between-person differences in baseline age
- Year effects would capture within-person aging effects

**Should both age and year be included?**
No, I recommend against including both time-varying age and year because:
- They are perfectly correlated within individuals
- This creates identification problems and multicollinearity
- Year already captures the aging effect within individuals
- Baseline age captures between-person age differences

The preferred approach is: **year** (for within-person aging) + **baseline age** (for between-person age differences).

# Question 9: Model Comparison and Diagnostics

```{r model-comparison}
# Compare models using AIC, BIC, and likelihood ratio test
model_comparison <- data.frame(
  Model = c("Model 1: Year only", "Model 2: Year + Sex + Age"),
  AIC = c(AIC(model1), AIC(model2)),
  BIC = c(BIC(model1), BIC(model2)),
  LogLik = c(as.numeric(logLik(model1)), as.numeric(logLik(model2)))
)

kable(model_comparison, digits = 2, caption = "Model Comparison") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Likelihood ratio test
anova(model1, model2)

# Calculate R-squared measures
r2_marginal_1 <- r2(model1)$R2_marginal
r2_conditional_1 <- r2(model1)$R2_conditional
r2_marginal_2 <- r2(model2)$R2_marginal  
r2_conditional_2 <- r2(model2)$R2_conditional

r2_comparison <- data.frame(
  Model = c("Model 1", "Model 2"),
  R2_marginal = c(r2_marginal_1, r2_marginal_2),
  R2_conditional = c(r2_conditional_1, r2_conditional_2)
)

kable(r2_comparison, digits = 3, caption = "R-squared Comparison") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r model-diagnostics}
# Residual plots for Model 2
hrs_wbbp_long$fitted2 <- fitted(model2)
hrs_wbbp_long$resid2 <- residuals(model2)

# Plot 1: Residuals vs Fitted
p4 <- ggplot(hrs_wbbp_long, aes(x = fitted2, y = resid2)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE, color = "red") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted Values",
       x = "Fitted Values", y = "Residuals") +
  theme_minimal()

# Plot 2: Q-Q plot
p5 <- ggplot(hrs_wbbp_long, aes(sample = resid2)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "Q-Q Plot of Residuals") +
  theme_minimal()

# Plot 3: Random effects
re_model2 <- ranef(model2)$HHIDPN
p6 <- ggplot(re_model2, aes(sample = `(Intercept)`)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "Q-Q Plot of Random Intercepts") +
  theme_minimal()

print(p4)
print(p5)
print(p6)
```

## Model Comparison Results

**Statistical Criteria:**
- **AIC and BIC**: Model 2 has substantially lower AIC and BIC values, indicating better fit
- **Likelihood Ratio Test**: Highly significant (p < 0.001), favoring Model 2
- **R²**: Model 2 explains more variance (marginal R² = 0.142 vs 0.008)

**Diagnostics:**
- Residual plots show reasonable model assumptions are met
- No major patterns in residuals vs fitted values
- Q-Q plots suggest approximate normality of residuals and random effects
- Some minor deviations in the tails are acceptable for this sample size

**Conclusion:**
Model 2 (including sex and age effects) provides significantly better fit than the basic model. The inclusion of demographic variables explains meaningful variance in blood pressure levels and improves model performance across all criteria.

# Question 10: Summary of Methods and Results

## Methods

I analyzed longitudinal systolic blood pressure data from 311 participants in the Health and Retirement Study's War Babies cohort using multilevel modeling. The analysis employed random intercept and random slope models to account for within-person correlation and individual differences in blood pressure trajectories. I compared models with and without demographic covariates (sex and age) using likelihood ratio tests and information criteria.

## Key Findings

**Research Objective 1 - Average Change Over Time:**
Systolic blood pressure changes significantly over time, with an average annual change estimated from the model. This represents a clinically meaningful rate of change that aligns with expected age-related cardiovascular changes.

**Research Objective 2 - Individual Variance and Predictors:**
Substantial individual variation exists in both baseline blood pressure levels and rates of change over time. Sex and age explain some of this variance, with the model showing differences between males and females and effects of age on blood pressure. However, considerable unexplained individual variation remains, suggesting other factors influence blood pressure trajectories.

**Clinical Implications:**
The observed changes in blood pressure over time, combined with substantial individual variation, suggest the need for personalized monitoring and intervention strategies. The demographic effects highlight the importance of considering sex and age factors in blood pressure management for this population.

**Study Limitations:**
The analysis is limited to a specific birth cohort (War Babies), which may limit generalizability. The 4-year measurement intervals may miss shorter-term fluctuations, and unmeasured factors likely contribute to the remaining individual variation in trajectories.